old snippets

### 2025-05-29 manuscript/manuscript.qmd
fix repo

# set up flextable for tables
set_flextable_defaults(
  font.family = "Times", 
  font.size = 10,
  font.color = "black",
  line_spacing = 1,
  padding.bottom = 1, 
  padding.top = 1,
  padding.left = 1,
  padding.right = 1
)
```

@@

#### Stefan Vermeent^1,2^, Anna-Lena Schubert^3^, & Willem E. Frankenhuis^2,4^

#### ^1^ Department of Psychology, Utrecht University, Utrecht, The Netherlands

#### ^2^ Max Planck Institute for the Study of Crime, Security, and Law, Freiburg, Germany

#### ^3^ Department of Psychology, University of Mainz, Mainz, Germany

#### ^4^ Evolutionary and Population Biology, Institute for Biodiversity and Ecosystem Dynamics, University of Amsterdam, Amsterdam, the Netherlands

@@

Stefan Vermeent served as lead for conceptualization, data curation, methodology, formal analysis, and writing the original draft. Anna-Lena Schubert served in a supporting role for conceptualization, formal analysis, methodology, and writing-review and editing. Willem E. Frankenhuis served in a supporting role for conceptualization, methodology, supervision, funding acquisition, and writing-review and editing.

The preregistration, analysis scripts, and study materials are available on the article's Github repository ([https://stefanvermeent.github.io/liss_ef_2024/](https://stefanvermeent.github.io/liss_ef_2024/)).
In this paper, we make use of data from the LISS panel (Longitudinal Internet studies for the Social Sciences) managed by the non-profit research institute Centerdata (Tilburg University, the Netherlands). 
All datasets are available to researchers in the LISS data archive, after signing a data use agreement ([https://www.dataarchive.lissdata.nl/study-units/view/1](https://www.dataarchive.lissdata.nl/study-units/view/1)).

Correspondence concerning this article should be addressed to Stefan Vermeent, Department of Psychology, Utrecht University, Heidelberglaan 1, 3584 CS Utrecht, The Netherlands. Email: p.c.s.vermeent@gmail.com / p.c.s.vermeent@uu.nl.

We have no conflict of interest to disclose.

**word count:** 9573

@@

Exposure to adversity may impair executive functioning (i.e., deficit frameworks), but could also enhance, or leave intact, specific EF abilities (i.e., adaptation frameworks).
Both frameworks often use raw performance (e.g., speed) to estimate EF ability. 
However, this approach (1) conflates different cognitive processes, and (2) generally does not distinguish specific EF abilities from processes that are shared across EF tasks, such as general processing speed.
Here, we integrate deficit and adaptation frameworks by building bridges with mathematical and cognitive psychology.
Specifically, we use cognitive modeling (Drift Diffusion Modeling) to isolate different cognitive processes: speed of information accumulation, response caution, and speed of stimulus encoding and response execution.
We then use structural equation modeling to investigate whether associations between adversity and cognitive processes are task-general or ability-specific.
We recruited 1061 participants from the Dutch LISS panel.
Participants completed a basic processing speed task, two inhibition tasks, and three attention-shifting tasks.
We measured exposure to threat and material deprivation in childhood and adulthood.
Exposure to threat (but not material deprivation) in adulthood was negatively associated with task-general processing speed.
After accounting for task-general processes, remaining variance was not related to either inhibition or attention-shifting ability.
Non-preregistered analyses showed that childhood exposure to material deprivation and threat were negatively associated with (1) general processing speed, and (2) task-specific information accumulation.
The latter reflected unique features of individual tasks, rather than specific EF abilities.
Taken together, these results suggest that adversity researchers overestimate associations between adversity and specific EF abilities when analyzing raw performance.

@@

Decades of research suggests that adversity impairs executive functioning abilities, such as inhibition and attention shifting.
These conclusions are often derived from analyzing raw performance, such as response times or accuracy.
We use cognitive modeling to reveal that adversity lowers general processing speed, but not specific executive functioning abilities.
This suggests that adversity research likely overestimates the extent to which adversity is associated with executive functioning abilities.

@@

@@

Recent research shows that raw performance on EF tasks does not necessarily capture specific EF abilities, for two related reasons.
First, raw performance on EF tasks is also influenced by processes other than the ability of interest, such as response caution and the speed of response execution [@hedge_2021; @ratcliff_2008; @stahl_2014].
Thus, slower (or faster) responses in people with more adversity exposure need not reflect EF ability, but could, for instance, reflect a tendency to respond more cautiously [@vermeent_2024a].
Second, raw performance on EF tasks is also influenced by general processes, such as basic speed of information processing [@frischkorn_2019; @loffler_2024; @weigard_2021].
Thus, lower performance on multiple EF tasks may reflect a difference in a general process, rather than specific EF abilities [@bignardi_2024; @vermeent_2024a; @young_2024].

@@

Cognitive modeling can distinguish between abilities and ability-irrelevant processes. 
Here, we focus on the Drift Diffusion Model [DDM\; @forstmann_2016; @ratcliff_1998; @ratcliff_2008; @wagenmakers_2009], which can be applied to many widely used EF tasks.
The DDM assumes that people accumulate evidence for one of two responses (e.g., left or right button) until they have acquired sufficient evidence to make a decision (see Figure 1).
When applied to trial-level response times and accuracy data, the DDM estimates four parameters that represent distinct cognitive processes [@voss_2004].
The *drift rate* reflects the rate at which people accumulate evidence for the correct response; thus, it measures the efficiency of information processing.
A higher drift rate leads to faster responses and higher accuracy.
When applied to EF tasks, the drift rate captures individual differences in specific EF abilities [although drift rate may also capture more general processes, see below\; @loffler_2024; @vermeent_2024a; @weigard_2021].
The *boundary separation* reflects the width between the two decision boundaries; it measures a person's response caution.
A higher boundary separation leads to slower responses and higher accuracy.
The *non-decision time* reflects a combination of preparation time (e.g., stimulus encoding) as well as execution time (e.g., time spent pressing the button).
A larger non-decision time leads to slower responses without a change in accuracy.
Finally, the *starting-point* allows for initial bias for one of two responses (e.g., happy or angry faces). 
In this paper, we will not consider the starting point, because people cannot have biases towards correct or incorrect responses.

@@

Recent studies show that a single general factor accounts for most of the variance in drift rates across multiple tasks [@lerche_2020; @loffler_2024; @weigard_2021].
This general factor appears to be largely stable across time [@schubert_2016; @weigard_2021].
It arises also in analyses of tasks thought to measure different abilities [@schmiedek_2007; @schmitz_2016].
Even if all tasks are designed to measure the same EF ability [@hedge_2021], variance in the general factor is fully explained by general processing speed instead of any specific EF ability [@loffler_2024].
These findings suggest that performance on standard EF tasks largely reflects basic speed of processing, rather than EF abilities.

Recent findings suggest that the effects of adversity are more general than specific.
For instance, youth with lower SES and more exposure to adversity show lower task-general raw performance scores across measures of EF, memory, and intelligence, while task-specific performance (after accounting for general variance) is often practically equivalent to zero [@bignardi_2024; @young_2024].
In addition, youth with more exposure to household threat showed lower task-general processing speed (based on drift rates), while differences in task-specific drift rates were practically equivalent to zero [@vermeent_2024a].
These findings suggest that previous research may have overestimated the associations between adversity and specific abilities.

A consequence of both sources of measurement impurity is that estimating specific EF abilities requires latent variable models including two or more measures of the same ability; yet, most studies in adversity research include only one task per EF ability.
This makes this work less suitable to investigate whether adversity is associated with specific EF abilities after accounting for general processes.
This is important for both deficit-based and adaptation-based interventions, which aim to remediate or leverage impaired or enhanced abilities, respectively. Moreover, while deficit frameworks predict impairments in both specific abilities and general processes [@vermeent_2024a], adaptation frameworks predict that specific types of adversity lead to adaptations in specific EF abilities.
For example, it has been proposed that unpredictable and dangerous environments may positively affect the ability to rapidly shift attention, promoting the detection of sudden threats and seizing fleeting opportunities [@frankenhuis_2020; @mittal_2015].
In the current research, therefore, we investigate the associations between adversity and two latent EF abilities—--inhibition and attention shifting—--after accounting for task-general processing speed.
We will use multiple tasks for each ability to obtain better estimates of specific abilities.

@@

Using a combination of DDM and structural equation modeling, we address three central questions.
First, what is the association of adversity exposure in adulthood with general processing speed that is contributing to performance across all cognitive tasks?
Second, what is the association of adversity exposure in adulthood with inhibition and attention-shifting abilities after accounting for general processing speed?
Third, what is the association of adversity exposure in adulthood with general and/or EF-specific response caution?
We focus on two types of adversity: material deprivation (a lack of access to material resources) and threat (the potential for harm imposed by others).
In previous research, both types of adversity were associated with performance on tasks used to measure EF [e.g., @fields_2021; @schafer_2022; @sheridan_2022; @vermeent_2024a; @young_2022].
In addition, both are central to recent dimensional models of adversity [@mclaughlin_2021; @sheridan_2014].

We preregistered predictions of deficit and adaptation frameworks.
Deficit frameworks predict negative associations between adversity and both the general speed of processing as well as specific abilities.
Adaptation frameworks predict intact or enhanced attention shifting ability, but not intact or enhanced inhibition ability.
We only interpret intact attention shifting ability as an adaptation if inhibition ability is also lower.
Intact performance on both abilities, however, would not provide sufficient evidence of adaptation, as it might instead suggest no association between adversity exposure and these abilities.
In line with previous work [@vermeent_2024a], we further predict that threat exposure is associated with higher task-general response caution.

@@

We tested a total of `r exclusions$sample$finish` participants from the Dutch Longitudinal Internet Studies for the Social Sciences (LISS) panel [@scherpenzeel_2011].
The LISS panel is an invitation-only, representative sample of the Dutch population consisting of approximately 7,500 individuals across 5,000 households.
LISS participants complete a yearly core battery of questionnaires about various domains of life, including one's financial situation over the past year.
In addition, LISS participants have the option to participate in further monthly studies on different topics.
We estimated power based on @kretzschmar_2019.
With $\alpha$ = 0.05 and assuming moderate reliability of our cognitive measures, we would have > 90% power to detect small effect sizes ($\beta$ = 0.1) with a sample size ranging between N = 730 and N = 980.
Therefore, we aimed for a total sample size of N = 1,000.

The current study took place between May and August 2024.
Participants were able to complete the cognitive tasks during two or more sessions to increase participation rates.
People were eligible for the study if they were between 18 and 55 years old, and had agreed to linking their LISS data to government microdata (not relevant here, but for a different study).
We excluded participants who did not have data on any cognitive task.
The final sample after exclusions consisted of `r exclusions$sample$final` participants (see Table 1).

```{r}
#| tab.id: table1
#| results: markup
table1 |> 
  flextable::autofit() |> 
  add_header_row(
    values = " ",
    colwidths = 2
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 1. "), "Descriptive statistics for the final sample."),
    part = "header"
  ) |> 
  border_remove() |> 
  border(i = 1, border.bottom = fp_border_default(), part = "header") |>  
  border(i = 2, border.top = fp_border_default(), part = "header") |>  
  border(i = 2, border.bottom = fp_border_default(), part = "header") |> 
  border(i = 14, border.bottom = fp_border_default(), part = "body") |> 
  autofit()
```

@@

Following our preregistration, we measured exposure to neighborhood violence in adulthood using two measures of perceived neighborhood crime, and one measure of crime victimization. 
Participants completed the Neighborhood Violence Scale [@frankenhuis_2018; @frankenhuis_2020].
This scale includes seven items about current perceived neighborhood threat (e.g., “Where I live, it is important to be able to defend yourself against physical harm”), on a scale of 1 ("Completely disagree") to 7 ("Completely agree"). 
These items were averaged and standardized.
We also included four items on perceived neighborhood threat from the LISS archive (six waves: <https://doi.org/10.17026/dans-zch-j8xt>).
Participants reported how frequently they: (1) “avoid certain areas in your place of residence because you perceive them as unsafe”, (2) “do not respond to a call at the door because you feel that it is unsafe”, (3) “leave valuable items at home to avoid theft or robbery in the street?”, (4) “make a detour, by car or on foot, to avoid unsafe areas?”.
The scale of these four items ranged from 1 (“(Almost) never”), 2 (“Sometimes”), to 3 (“Often”).
We intended to include these items again in the current study, but due to an oversight this did not happen.
This meant that we did not have data on these items for participants who never participated in any of the six waves described above (missing N = `r exclusions$sample$missing_crime_vict`). 
We decided to conduct all analyses involving threat based only on this subset of the data (N = `r exclusions$sample$final - exclusions$sample$missing_crime_vict`), because *a priori* power remained above 80% [@kretzschmar_2019].
We summed the items within each wave, and then calculated an average across waves for which participants had data.

@@

The correlations between the two measures of perceived neighborhood crime and crime victimization were low (see Table 2).
Therefore, following our preregistration, we used Principal Component Analysis to extract the first principal component score.
This score accounted for 23 % of the variance in the three measures.

@@

```{r}
#| tab.id: table2
#| results: markup
table2 |> 
  flextable() |> 
  flextable::set_header_labels(Variable = "") |> 
  flextable::autofit() |> 
  border_remove() %>% 
 # border(i = 1, border.top = fp_border_default(), part = "header") %>% 
  border(i = 1, border.bottom = fp_border_default(), part = "header") %>% 
  border(i = 18, border.bottom = fp_border_default(), part = "body") %>% 
  add_header_row(
    values = " ",
    colwidths = 13
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 2. "), "Spearman correlations between the main independent variables and covariates."),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 13
  ) |> 
  autofit()
```

@@

*Global-local Task.* This task measures the ability to shift attention between tasks.
We adapted the stimuli from @huizinga_2010.
On each trial, participants saw a large square or rectangle composed of 16 small squares or rectangles.
The stimulus was flanked on both side by a drawing of an elephant or mouse, which was presented 1,000 ms prior to the appearance of the stimulus.
If the stimulus was flanked by the image of an elephant (50% of trials), participants had to indicate whether the global image was a square or rectangle.
If the stimulus was flanked by the image of a mouse (50% of trials), participants had to indicate whether the local images were squares or rectangles.
On 50 % of the trials, the rule was the same as on the preceding trial (repeat trials).
On the other half, the rule was different than on the preceding trial (switch trials).
Finally, the stimuli were congruent on 50% of the trials (e.g., large square consisting of small squares) and incongruent on the other half (e.g., large square consisting of small rectangles).
Congruency, task rule (switch vs. repeat), or focus (global vs. local) where never repeated more than three times in a row.
Participants first completed eight practice trials, followed by two test blocks of 32 trials each, for a total of 64 trials.

@@

Participants reported their state anxiety after each cognitive task; thus, six times in total.
We measured state anxiety using the shortened version of the State-Trait Anxiety Inventory [@bij_2003; @marteau_1992], which asks participants how calm, tense, upset, relaxed, content, and worried they currently feel, on a scale of 1 ("not al all") to 4 ("very much").
We recoded (if necessary) and then summed the answers with higher scores reflecting more state anxiety.

@@

Participants reported the level of environmental noise after each cognitive task; thus, six times in total.
We measured environmental noise using a single item, rated on a scale of 1 to 5: "How much noise was there in your environment during the game?"

@@

The set of potential confounders consisted of (1) age [@salthouse_2016; @salthouse_2019; @starns_2010]; (2) education [@hofmarcher_2021]; (3) sex [@ning_2023]; (4) childhood adversity exposure (material deprivation and threat combined) [@bos_2009; @goodman_2019]; and (5) potential causal relations between recent material deprivation and recent threat [@bywaters_2016; @lacey_2022; @ning_2023].
We made these decisions based on previous literature, or, in cases of doubt, by statistically testing support for specific relations using data from a previous LISS study [@vermeent_2024c].
Specifically, following our preregistration, we used the *dagitty* R package [@ankan_2021; @textor_2016] to test whether independencies between specific variables implied by the Directed Acyclic Graph (conditional on the other causal pathways) were supported by these data (see the preregistration for more details).

>>>>

# set up flextable for tables
set_flextable_defaults(
  font.family = "Times", 
  font.size = 10,
  font.color = "black",
  line_spacing = 1,
  padding.bottom = 1, 
  padding.top = 1,
  padding.left = 1,
  padding.right = 1,
  opts_word = list(split = FALSE, keep_with_next = TRUE)
)
```

#### **Adversity is associated with lower general processing speed rather than executive functioning**

@@

#### Stefan Vermeent^1,2,3^, Anna-Lena Schubert^4^, & Willem E. Frankenhuis^1,3^

#### ^1^ Evolutionary and Population Biology, Institute for Biodiversity and Ecosystem Dynamics, University of Amsterdam, Amsterdam, the Netherlands

#### ^2^ Department of Psychology, Utrecht University, Utrecht, The Netherlands

#### ^3^ Max Planck Institute for the Study of Crime, Security, and Law, Freiburg, Germany

#### ^4^ Department of Psychology, University of Mainz, Mainz, Germany

@@

This study was approved by the Ethics Review Board of the Faculty of Social & Behavioural Sciences of Utrecht University (FETC20-490) and the Ethics committee for research in the Sciences and Life Sciences of the University of Amsterdam (FNWI-41_2023).

All scripts and materials needed to reproduce the findings are available on the article's Github repository ([https://stefanvermeent.github.io/liss_ef_2024/](https://stefanvermeent.github.io/liss_ef_2024/)).
In this paper, we make use of data from the LISS panel (Longitudinal Internet studies for the Social Sciences) managed by the non-profit research institute Centerdata (Tilburg University, the Netherlands). 
All datasets are available in the LISS data archive ([https://www.dataarchive.lissdata.nl/study-units/view/1](https://www.dataarchive.lissdata.nl/study-units/view/1)). 
Researchers who want to access the data are required to sign a statement confirming that information about individual persons, households, etc., will not be released to others (go to [https://statements.centerdata.nl](https://statements.centerdata.nl) for more information).
The data and ideas in this manuscript were not presented elsewhere.

SV served as lead for conceptualization, formal analysis, funding acquisition, investigation, methodology, and writing - original draft.
AS served in a supporting role for supervision, conceptualization, methodology, formal analysis, and writing - review & editing.
WF served as lead for supervision, and in a supporting role for conceptualization, methodology, formal analysis, and writing - review & editing.

The authors declare no conflicts of interest.

Correspondence concerning this article should be addressed to Stefan Vermeent, Evolutionary and Population Biology, Institute for Biodiversity and Ecosystem Dynamics, University of Amsterdam, Science Park 904, 1098 XH Amsterdam, The Netherlands. Email: p.c.s.vermeent@gmail.com.

\pagebreak

@@

Exposure to adversity may impair executive functioning (i.e., deficit frameworks), but could also enhance, or leave intact, specific EF abilities (i.e., adaptation frameworks).
Both frameworks often use raw performance (e.g., speed) to estimate EF ability. 
However, this approach (1) conflates different cognitive processes, and (2) generally does not distinguish specific EF abilities from processes that are shared across EF tasks, such as general processing speed.
Here, we integrate deficit and adaptation frameworks by building bridges with mathematical and cognitive psychology.
Specifically, we use cognitive modeling (Drift Diffusion Modeling) to isolate different cognitive processes: speed of information accumulation, response caution, and speed of stimulus encoding and response execution.
We then use structural equation modeling to investigate whether associations between adversity and cognitive processes are task-general or ability-specific.
We recruited 1061 participants from the Dutch LISS panel.
Participants completed a basic processing speed task, two inhibition tasks, and three attention-shifting tasks.
We measured self-reported exposure to threat and material deprivation in childhood and adulthood.
Exposure to threat (but not deprivation) in adulthood was negatively associated with task-general processing speed.
After accounting for task-general processes, remaining variance was not related to either inhibition or attention-shifting ability.
Non-preregistered analyses showed that childhood exposure to deprivation and threat were negatively associated with (1) general processing speed, and (2) task-specific information accumulation.
The latter reflected unique features of individual tasks, rather than specific EF abilities.
Taken together, these results suggest that adversity researchers may overestimate associations between adversity and specific EF abilities when analyzing raw performance.

@@

Decades of research suggests that adversity impairs executive functioning abilities, such as inhibition and attention shifting.
These conclusions are often derived from analyzing raw performance, such as response times or accuracy.
We use cognitive modeling to reveal that self-reported adversity is associated with general processing speed, but not specific executive functioning abilities.
This suggests that adversity research may overestimate the extent to which adversity is associated with executive functioning abilities.

@@

Most research in this field has focused on cognitive abilities in childhood, or associations between early-life adversity and cognitive abilities up to early adulthood. 
However, adaptations to early adversity may persist into adulthood, and people may also be able to develop cognitive adaptations in response to adversity exposure later in life. 
First, there is evidence that early-life adversity is associated with both cognitive deficits and adaptations in middle and late adulthood. 
Women who reported a history of sexual abuse before the age of 18 showed better delayed recall and better visual reasoning at age 50-64 [@feeney_2013]
Similarly, self-reported material deprivation in early adulthood was associated with better delayed recall and verbal fluency in middle adulthood [@kunzi_2023]. 
In contrast, higher levels of self-reported childhood abuse have also been linked to lower cognitive performance in middle- to late adulthood (@roberts_2022). 
Second, people may also develop cognitive adaptations in response to adversity experienced later in life. 
People who reported higher levels of material deprivation in middle adulthood showed a smaller decline in delayed recall compared to people with lower levels of material deprivation [@kunzi_2023]. 
These findings highlight the importance of studying the effects of adversity across the lifespan.

@@

Recent research shows that raw performance on EF tasks does not necessarily capture specific EF abilities, for two related reasons.
First, raw performance on EF tasks is also influenced by processes other than the ability of interest, such as response caution and the speed of response execution [@hedge_2022; @ratcliff_2008; @stahl_2014].
Thus, slower (or faster) responses in people with more adversity exposure need not reflect EF ability, but could, for instance, reflect a tendency to respond more cautiously [@vermeent_2024a].
Second, raw performance on EF tasks is also influenced by general processes, such as basic speed of information processing [@frischkorn_2019; @loffler_2024; @weigard_2021].
Thus, lower performance on multiple EF tasks may reflect a difference in a general process, rather than specific EF abilities [@bignardi_2024; @vermeent_2024a; @young_2024].

@@

Cognitive modeling can distinguish between abilities and ability-irrelevant processes. 
Here, we focus on the Drift Diffusion Model [DDM\; @forstmann_2016; @ratcliff_1998; @ratcliff_2008; @wagenmakers_2009], which can be applied to many widely used EF tasks.
The DDM assumes that people accumulate evidence for one of two responses (e.g., left or right button) until they have acquired sufficient evidence to make a decision (see Figure 1).
When applied to trial-level response times and accuracy data, the DDM estimates three parameters that represent distinct cognitive processes [@voss_2004].
The *drift rate* reflects the rate at which people accumulate evidence for the correct response; thus, it measures the efficiency of information processing.
A higher drift rate leads to faster responses and higher accuracy.
When applied to EF tasks, the drift rate captures individual differences in specific EF abilities [although drift rate may also capture more general processes, see below\; @loffler_2024; @vermeent_2024a; @weigard_2021].
The *boundary separation* reflects the width between the two decision boundaries; it measures a person's response caution.
A higher boundary separation leads to slower responses and higher accuracy.
The *non-decision time* reflects a combination of preparation time (e.g., stimulus encoding) as well as execution time (e.g., time spent pressing the button).
A larger non-decision time leads to slower responses without a change in accuracy.
Finally, the *starting-point* allows for initial bias for one of two responses (e.g., happy or angry faces). 
In this paper, we will not consider the starting point, because people cannot have biases towards correct or incorrect responses.

How do the DDM parameters relate to executive functioning ability?
Not all EF tasks require a person to accumulate evidence under time pressure, and some research suggests that EF measurement may be improved by removing speed requirements altogether [e.g., @draheim_2021b].
However, many common EF tasks do depend on evidence accumulation.
For instance, the Flanker Task requires accumulating evidence for the arrow direction (while not letting the flanking arrows contaminate this evidence), and attention-shifting tasks require accumulating evidence for the goal that is currently relevant.
Given the popularity of these tasks in adversity research, associations between adversity exposure and EF have largely been studied using response times. 
This raises the possibility that these associations are caused by ability-irrelevant processes rather than EF abilities. 
If performance differences really reflect EF abilities, they should be reflected in drift rates, not boundary separations or non-decision times.

@@

Recent studies show that a single general factor accounts for most of the variance in drift rates across multiple EF tasks [@lerche_2020; @loffler_2024; @weigard_2021].
This general factor appears to be largely stable across test sessions [@schubert_2016; @weigard_2021], and has been interpreted as reflecting general processing speed [@hedge_2022; @loffler_2024].
Although general processing speed exerts a strong influence on EF task performance, it is conceptually distinct from EF abilities. 
Processing speed refers to the rate at which elementary cognitive operations are completed, whereas executive functions involve higher-order cognitive processes that support goal-directed behavior, particularly under conditions of distraction or interference [@mashburn_2024; @vonbastian_2020]. 
Thus, to isolate specific EF abilities from raw task performance, it is essential to account for the influence of general processing speed.
This approach has been used in mathematical and cognitive psychology, but hardly in research on the impact of adversity on EF abilities.

We are aware of three studies that estimated the associations between adversity and specific cognitive abilities after partialing out general processing.
These studies suggest that the effects of adversity are more general than specific.
Two studies showed that youth with lower socioeconomic status and more exposure to adversity show lower task-general raw performance scores across measures of EF, memory, and intelligence, while task-specific performance (after accounting for general variance) is often practically equivalent to zero [@bignardi_2024; @young_2024].
A third study showed that youth with more exposure to household threat showed lower task-general processing speed (based on drift rates), while differences in task-specific drift rates were practically equivalent to zero [@vermeent_2024a].
Jointly, these findings suggest that previous research may have overestimated the associations between adversity and specific abilities.

A consequence of measurement impurity is that estimating specific EF abilities requires latent variable models including two or more measures of the same ability. Yet, most studies in adversity research include only one task per EF ability.
This makes this work less suitable to investigate whether adversity is associated with specific EF abilities after accounting for general processes.
This is important for both deficit-based and adaptation-based interventions, which aim to remediate or leverage impaired or enhanced abilities, respectively. 
Moreover, while deficit frameworks predict impairments in both specific abilities and general processes [@vermeent_2024a], adaptation frameworks predict that specific types of adversity lead to adaptations in specific EF abilities only.
For example, it has been proposed that unpredictable and dangerous environments may positively affect the ability to rapidly shift attention, promoting the detection of sudden threats and seizing fleeting opportunities [@frankenhuis_2020; @mittal_2015].
In the current research, therefore, we investigate the associations between adversity and two latent EF abilities—--inhibition and attention shifting—--after accounting for task-general processing speed.
We will use multiple tasks for each ability to obtain better estimates of specific abilities.

@@

Using a combination of DDM and structural equation modeling, we address three central questions.
First, what is the association of self-reported adversity exposure in adulthood with general processing speed that is contributing to performance across all cognitive tasks?
Second, what is the association of self-reported adversity exposure in adulthood with inhibition and attention-shifting abilities after accounting for general processing speed?
Third, what is the association of self-reported adversity exposure in adulthood with general and/or EF-specific response caution?
We focus on two types of adversity: material deprivation (a lack of access to material resources) and environmental threat (the potential for harm imposed by others).
In previous research, both types of adversity were associated with performance on tasks used to measure EF [e.g., @fields_2021; @schafer_2022; @sheridan_2022; @vermeent_2024a; @young_2022].
In addition, both are central to recent dimensional models of adversity [@mclaughlin_2021; @sheridan_2014].
In contrast to specificity models (which distinguish between separate distinct adversities, such as verbal or physical abuse) and cumulative-risk models (which add-up discrete exposures to various adverse experiences), dimensional models of adversity aim to identify dimensions of adversity, consisting of different types of adversity with similar features [@mclaughlin_2021].

We preregistered predictions of deficit and adaptation frameworks.
From a deficit framework, we predict negative associations between adversity and both the general speed of processing as well as specific abilities. 
From an adaptation framework, we have three predictions.
First, we predict a positive association between threat exposure and attention-shifting ability.
This follows from the hypothesis that violence exposure might lead to broader, more vigilant monitoring of information, which could help to detect threats early on [@frankenhuis_2016; @vermeent_2025].
We only interpret intact attention shifting ability in association with threat as an adaptation if inhibition ability is also lower. 
Intact performance on both abilities, however, would not provide sufficient evidence of adaptation, as it might instead suggest no association between adversity exposure and these abilities. 
Second, we expect that broader, more vigilant monitoring of information would hurt performance on inhibition tasks, as it would result in increased processing of distractions.
Therefore, we do not predict a positive association between threat and inhibition ability.
Third, although one could argue that material deprivation might also increase monitoring (of potential resources), no previous work that we are aware of has found a positive association between deprivation and attention shifting [@mittal_2015; @young_2022].
Thus, we do not predict a positive association between material deprivation and attention-shifting ability (nor inhibition ability).
Finally, in line with previous work [@vermeent_2024a], we further predict that threat exposure is associated with higher task-general response caution.

@@

We tested a total of `r exclusions$sample$finish` participants from the Dutch Longitudinal Internet Studies for the Social Sciences (LISS) panel [@scherpenzeel_2011].
The LISS panel is an invitation-only, representative sample of the Dutch population consisting of approximately 7,500 individuals across 5,000 households.
LISS participants complete a yearly core battery of questionnaires about various domains of life, including one's financial situation over the past year.
In addition, LISS participants have the option to participate in further monthly studies on different topics.
Each month, participants have the opportunity to update their demographic information (e.g., sex, education level, household composition).
We estimated power based on @kretzschmar_2019.
With $\alpha$ = 0.05 and assuming moderate reliability of our cognitive measures, we would have > 90% power to detect small effect sizes ($\beta$ = 0.1) with a sample size ranging between N = 730 and N = 980.
Therefore, we aimed for a total sample size of N = 1,000.

The current study took place between May and August 2024.
Participants were able to complete the cognitive tasks during two or more sessions to increase participation rates.
People were eligible for the study if they were between 16 and 55 years old, and had agreed to linking their LISS data to government microdata (not relevant here, but for a different study).
We excluded participants who did not have data on any cognitive task.
The final sample after exclusions consisted of `r exclusions$sample$final` participants (see Table 1).

@@

Following our preregistration, we measured exposure to neighborhood violence in adulthood using three measures: two measures of perceived neighborhood crime, and one measure of crime victimization. 
Participants completed the Neighborhood Violence Scale [@frankenhuis_2018; @frankenhuis_2020].
This scale includes seven items about current perceived neighborhood threat (e.g., “Where I live, it is important to be able to defend yourself against physical harm”), on a scale of 1 ("Completely disagree") to 7 ("Completely agree"). 
These items were averaged and standardized.
We also included four items on perceived neighborhood threat from the LISS archive (six waves: <https://doi.org/10.17026/dans-zch-j8xt>).
Participants reported how frequently they: (1) “avoid certain areas in your place of residence because you perceive them as unsafe”, (2) “do not respond to a call at the door because you feel that it is unsafe”, (3) “leave valuable items at home to avoid theft or robbery in the street?”, (4) “make a detour, by car or on foot, to avoid unsafe areas?”.
The scale of these four items ranged from 1 (“(Almost) never”), 2 (“Sometimes”), to 3 (“Often”).
We intended to include these items again in the current study, but due to an oversight this did not happen.
This meant that we did not have data on these items for participants who never participated in any of the six waves described above (missing N = `r exclusions$sample$missing_crime_vict`). 
We decided to conduct all analyses involving threat based only on this subset of the data (N = `r exclusions$sample$final - exclusions$sample$missing_crime_vict`), because *a priori* power remained above 80% [@kretzschmar_2019].
We summed the items within each wave, and then calculated an average across all the waves for which we had a participant's data.

@@

The correlations between the two measures of perceived neighborhood crime and crime victimization were low (see Table 2).
Therefore, following our preregistration, we used Principal Component Analysis to extract the first principal component score.
This score accounted for 23% of the variance in the three measures.

@@

### Childhood adversity

Participants reported their exposure to childhood deprivation (prior to age 18) in the new data collection using four items: (1) "My family had enough money to afford the kind of home we all needed"; (2) "My family had enough money to afford the kind of clothing we all needed"; (3) "My family had enough money to afford the kind of food that we all needed"; (4) "My family had enough money to afford the kind of medical care that we all needed", on a scale from 1 (completely not agree) to 7 (completely agree). 
This measure differed from those used to measure exposure to deprivation in adulthood. 
In addition, participants reported their exposure to childhood threat (prior to age 18) in the new data collection using the Neighborhood Violence Scale ("Crime was common in the neighborhood where I grew up") , on a scale from 1 (completely not agree) to 7 (completely agree) [@frankenhuis_2018; @frankenhuis_2020]. 
This measure was also used to measure threat exposure in adulthood.

When used as covariates in the main analyses (see Confounds section), the measures of childhood deprivation and childhood threat were averaged together into a single measure of childhood adversity. 
When used as main predictors in the exploratory analyses that we conducted, they were treated separately.

@@

*Global-local Task.* This task measures the ability to shift attention between tasks.
We adapted the stimuli from @huizinga_2010.
On each trial, participants saw a large square or rectangle composed of 16 small squares or rectangles.
The stimulus was flanked on both side by a drawing of an elephant or mouse, which was presented 1,000 ms prior to the appearance of the stimulus.
If the stimulus was flanked by the image of an elephant (50% of trials), participants had to indicate whether the global image was a square or rectangle.
If the stimulus was flanked by the image of a mouse (50% of trials), participants had to indicate whether the local images were squares or rectangles.
On 50% of the trials, the rule was the same as on the preceding trial (repeat trials).
On the other half, the rule was different than on the preceding trial (switch trials).
Finally, the stimuli were congruent on 50% of the trials (e.g., large square consisting of small squares) and incongruent on the other half (e.g., large square consisting of small rectangles).
Congruency, task rule (switch vs. repeat), or focus (global vs. local) where never repeated more than three times in a row.
Participants first completed eight practice trials, followed by two test blocks of 32 trials each, for a total of 64 trials.

@@

Participants reported their state anxiety after each cognitive task; thus, six times in total.
We measured state anxiety using the shortened version of the State-Trait Anxiety Inventory [@bij_2003; @marteau_1992], which asks participants how calm, tense, upset, relaxed, content, and worried they currently feel, on a scale of 1 ("not al all") to 4 ("very much").
We recoded (if necessary) and then summed the answers, with higher scores reflecting more state anxiety.

@@

Participants reported the level of environmental noise after each cognitive task; thus, six times in total.
We measured environmental noise using a single item, rated on a scale of 1 ("very little noise") to 5 ("a lot of noise"): "How much noise was there in your environment during the game?".

@@

The set of potential confounders consisted of (1) age [@salthouse_2016; @salthouse_2019; @starns_2010]; (2) education [@hofmarcher_2021]; (3) sex [@ning_2023]; (4) childhood adversity exposure (material deprivation and threat combined) [@bos_2009; @goodman_2019]; and (5) potential causal relations between recent material deprivation and recent threat [@bywaters_2016; @lacey_2022; @ning_2023].
We made these decisions based on previous literature, or, in cases of doubt, by statistically testing support for specific relations using data from a previous LISS study [@vermeent_2025].
Specifically, following our preregistration, we used the *dagitty* R package [@ankan_2021; @textor_2016] to test whether independencies between specific variables implied by the Directed Acyclic Graph (conditional on the other causal pathways) were supported by these data (see the preregistration for more details).

Figure 2 depicts the final preregistered Directed Acyclic Graph for material deprivation and threat exposure in adulthood.
For threat exposure in adulthood, our statistical model controlled for age, sex, childhood adversity, and recent material deprivation.
We included material deprivation in adulthood based on studies showing that material deprivation tends to precede exposure to adversities such as threat [@bywaters_2016; @lacey_2022; @ning_2023].
We did not control for education given a lack of theoretical and statistical support for this effect (see preregistration).
To explore the effect of our assumption that material deprivation tends to precede threat, we also present a secondary model excluding material deprivation in adulthood as a confound.
As preregistered, we will not base our main conclusions on this secondary model.
